Using Deep Learning to Identify Patients with Cognitive Impairment in Electronic Health Records

### 2021-11-02 Submission/Paper/ML4H/Paper.tex
edits to make it camera ready since it has been accepted

 %\title{Title of the Article\titletag{\thanks{Some footnote}}}

% \author{Tanish Tyagi \nametag{\thanks{Authors contributed equally}\SUP{1}},
% Colin G. Magdamo\nametag{\footnotemark[1]\SUP{1}}, 
% Ayush Noori \SUP{1},
% Mayuresh Deodhar \SUP{1},
% Zhuoqiao Hong \SUP{1},
% Alberto Serrano-Pozo \SUP{1},
% Dimitry Prokopenko \SUP{1},
% Rudy E. Tanzi \SUP{2},
% Deborah Blacker \SUP{1},
% Bradley T. Hyman \SUP{1},
% Shibani S. Mukerji \SUP{1},
% M. Brandon Westover \SUP{1},
% Sudeshna Das\SUP{1},
% \centering \Email{
% \\[\bigskipamount] 
% \SUP{1}\{ttyagi, 
% cmagdamo,
% anoori1,
% mdeodhar,
% zhong1,
% aserrano1,
% dprokopenko,
% dblacker,
% bhyman,
% smukerji,
% mwestover,
% sdas5\}
% @mgh.harvard.edu}
% \centering \Email{
% \\[\bigskipamount] 
% \SUP{2}\{tanzi\}@helix.mgh.harvard.edu
% }
% 
% \begin{center}\addr Massachusetts General Hospital, Boston, MA\end{center}
% }

@@

\section{Introduction}
\label{sec:intro} Dementia is the most common neurodegenerative disease affecting older adults, progressing from mild cognitive impairment (MCI) to mild, moderate, and severe dementia. Dementia is under-diagnosed: dementia is not formally diagnosed or coded in claims data for over 50\% of older adults living with probable dementia. Often, a diagnosis is given once patient has reached moderate dementia, and irreversible damage has already been done to the brain.  
% https://www.dementiacarecentral.com/aboutdementia/facts/stages/ %
The early detection of the first signs of cognitive impairment, however, is important for improving clinical outcomes and patient management. Tools that can efficiently and effectively analyze medical records for warning signs of dementia and recommend patients for follow up with a specialist can be critical to obtaining an early diagnosis for dementia. Such a tool could also be useful in recruiting into clinical trials as well as a variety of research studies ranging from in-silico trials for drug repurposing to evaluating how policies and programs meet the needs of patients and caregivers.  We aim to use NLP to detect signs of cognitive impairment from unstructured clinician notes by using deep learning techniques. We apply our deep learning algorithm to patients in Mass General Brigham (MGB) Healthcare who have genotype data available from the MGB BioBank. An overview of our project can be found in Appendix \ref{app:overview}.

%% Table 1 - Demographics of Dataset %%
\begin{table*}[hbtp]
\centering 
\floatconts
{tab:table1}
{\caption{Demographics of Data}} 
    {
        \begin{tabular}{lccc}
        \toprule
        \bfseries Characteristic & \bfseries (N = 16428) \\
        \midrule
        \textbf{Age (years) mean (SD)} & \fseries 73.01 (7.96) \\ 
        \textbf{Gender Male, \emph n (\%)}  & \fseries 8740 (53.2)\\ 
        \textbf{Race, \emph n (\%)} \\ 
            \hspace{10mm} White & 14896 (90.7) \\
            \hspace{10mm} Other/Not Recorded & 608 (3.7) \\
            \hspace{10mm} Black & 570 (3.5) \\
            \hspace{10mm} Hispanic & 170 (1.0) \\
            \hspace{10mm} Asian & 168 (1.0) \\
            \hspace{10mm} Indigenous & 16 (0.01) \\
        \textbf{APOE Genotype, \emph n (\%)} \\ 
            \hspace{10mm} {APOE ${\bm{\varepsilon}}$2 & 2028 (12.3)} \\
            \hspace{10mm} {APOE ${\bm{\varepsilon}}$3 & 10177 (62.0)} \\
            \hspace{10mm} {APOE ${\bm{\varepsilon}}$4 & 4223 (25.7)} \\
        \textbf{Average Speciality Visits (SD)} & 1.67 (4.6) \\ 
        \textbf{Average PCP Encounters (SD)} &  5.25 (5.63) \\ 
        % \textbf{Encounters AVG (SD)} & 14.55 (23.2) \\
        % \textbf{Keyword Matches AVG (SD)} & 26 (54.03) \\
        \bottomrule
        \end{tabular}
    }
\end{table*}

@@

@@

ClinicalBERT, with its more complex architecture, was able to leverage the context of the keyword matches within the sequences and overcome these issues. The fine-tuned ClinicalBERT model achieved an AUC of 0.98 and substantially improved accuracy to 0.93 (specificity of 0.96, sensitivity of 0.91, micro F1 of 0.93, macro F1 of 0.92, and weighted F1 of 0.93).

>>>>

 %\title{Title of the Article\titletag{\thanks{Some footnote}}}
% figure out later

% \author{test \nametag{\thanks{Authors contributed equally}},
% \centering \Email{ 
% \\[\bigskipamount]  
% TEST, }
% }

% ===============================================================

\author{Tanish Tyagi \nametag{\thanks{Authors contributed equally} \SUP{1}},
Colin G. Magdamo\nametag{\footnotemark[1]\SUP{1}}, 
Ayush Noori \SUP{1},
Zhaozhi Li \SUP{1},
Xiao Liu \SUP{1},
Mayuresh Deodhar \SUP{1},
Zhuoqiao Hong \SUP{1},
Wendong Ge \SUP{1},
Alberto Serrano-Pozo \SUP{1},
Sahar Zafar \SUP{1}, 
Nicole Benson \SUP{1},
Lidia Moura \SUP{1},
John Hsu \SUP{1},
Dimitry Prokopenko \SUP{1},
Rudy E. Tanzi \SUP{1},
Bradley T. Hyman \SUP{1},
Deborah Blacker \SUP{1},
Shibani S. Mukerji \SUP{1},
M. Brandon Westover \SUP{1},
Sudeshna Das\SUP{1}
\centering \Email{
\\[\bigskipamount] 
\SUP{1}\{ttyagi, 
cmagdamo,
anoori1,
zli39,
xliu61,
mdeodhar,
zhong1,
wendong.ge,
aserrano1,
sfzafar, 
nbenson, 
lidia.moura, 
john.hsu,
dprokopenko,
dblacker,
rtanzi,
bhyman,
smukerji,
mwestover,
sdas5\}
@mgh.harvard.edu}
\centering
\begin{center}\addr Massachusetts General Hospital, Boston, MA\end{center}
}

@@

\section{Introduction}
\label{sec:intro} Dementia is the most common neurodegenerative disease affecting older adults, progressing from mild cognitive impairment (MCI) to mild, moderate, and severe dementia. Dementia is under-diagnosed: dementia is not formally diagnosed or coded in claims data for over 50\% of older adults living with probable dementia. Often, a diagnosis is given once patient has reached moderate dementia, and irreversible damage has already been done to the brain.  
% https://www.dementiacarecentral.com/aboutdementia/facts/stages/ %
The early detection of the first signs of cognitive impairment, however, is important for improving clinical outcomes and patient management. Tools that can efficiently and effectively analyze medical records for warning signs of dementia and recommend patients for follow up with a specialist can be critical to obtaining an early diagnosis for dementia. We aim to use NLP to detect signs of cognitive impairment from unstructured clinician notes by using deep learning techniques. We apply our deep learning algorithm to patients in Mass General Brigham (MGB) Healthcare who have genotype data available from the MGB BioBank. An overview of our project can be found in Appendix \ref{app:overview}.

@@

\begin{table}[htb]
\resizebox{\columnwidth}{!}{%
\begin{tabular}{lccc}
        \toprule
        \bfseries Characteristic & \bfseries (N = 16428) \\
        \midrule
        \textbf{Age (years) mean (SD)} & \fseries 73.01 (7.96) \\ 
        \textbf{Gender Male, \emph n (\%)}  & \fseries 8740 (53.2)\\ 
        \textbf{Race, \emph n (\%)} \\ 
            \hspace{10mm} White & 14896 (90.7) \\
            \hspace{10mm} Other/Not Recorded & 608 (3.7) \\
            \hspace{10mm} Black & 570 (3.5) \\
            \hspace{10mm} Hispanic & 170 (1.0) \\
            \hspace{10mm} Asian & 168 (1.0) \\
            \hspace{10mm} Indigenous & 16 (0.01) \\
        \textbf{APOE Genotype, \emph n (\%)} \\ 
            \hspace{10mm} {APOE ${\bm{\varepsilon}}$2 & 2028 (12.3)} \\
            \hspace{10mm} {APOE ${\bm{\varepsilon}}$3 & 10177 (62.0)} \\
            \hspace{10mm} {APOE ${\bm{\varepsilon}}$4 & 4223 (25.7)} \\
        \textbf{Average Speciality Visits (SD)} & 1.67 (4.6) \\ 
        \textbf{Average PCP Encounters (SD)} &  5.25 (5.63) \\ 
        % \textbf{Encounters AVG (SD)} & 14.55 (23.2) \\
        % \textbf{Keyword Matches AVG (SD)} & 26 (54.03) \\
        \bottomrule
\end{tabular}
}
\end{table}

@@

ClinicalBERT, with its more complex architecture, was able to leverage the context of the keyword matches within the sequences and overcome these issues. Indeed, using a small dataset of manually annotated sequences (n=150) which did not match an always-pattern, the ClinicalBERT embeddings were able to able accurately discriminate between all three classes (see Appendix). The fine-tuned ClinicalBERT model achieved an AUC of 0.98 and substantially improved accuracy to 0.93 (specificity of 0.96, sensitivity of 0.91, micro F1 of 0.93, macro F1 of 0.92, and weighted F1 of 0.93).
